---
title: "Homework 2"
author: "Lilit Ivanyan"
date: "2025-02-20"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
### Part 3

The “lung_cancer_prediction_dataset.csv” provides valuable insights into lung cancer cases, risk
factors, smoking trends, and healthcare access across 25 of the world's most populated
countries. It includes 220,632 individuals with details on their age, gender, smoking history,
cancer diagnosis, environmental exposure, and survival rates. The dataset is useful for medical
research, predictive modeling, and policymaking to understand lung cancer patterns globally.

The “global_air_pollution_dataset.csv” provides information about the contamination of the
indoor or outdoor environment by any chemical, physical or biological agent that modifies the
natural characteristics of the atmosphere. It includes 23,463 rows with data presented in the
following columns: Country, City, AQI Value, AQI Category, CO AQI Category, Ozone AQI Value,
Ozone AQI Category, NO2 AQI Value, NO2 AQI Category, PM2.5 AQI Value, PM2.5 AQI Category.

```{r}
library(ggplot2)
library(dplyr)
library(viridis)

# Load the data
lung_cancer_data <- read.csv('lung_cancer_prediction_dataset.csv')
air_pollution_data <- read.csv('global_air_pollution_dataset.csv')
```

1. Create a Boxplot of Lung Cancer Deaths Distribution. 
```{r}
boxplot_lung_cancer <- ggplot(lung_cancer_data, aes(y = Annual_Lung_Cancer_Deaths)) +
  geom_boxplot(fill = "lightblue", color = "blue") +
  labs(title = 'Boxplot of Annual Lung Cancer Deaths Distribution', y = 'Number of Annual Deaths')

print(boxplot_lung_cancer)
```

2. Create a Histogram of PM2.5 AQI Values. 
```{r}

histogram <- ggplot(air_pollution_data, aes(x = AQI_Value)) +
  geom_histogram(bins = 30, fill="blue", alpha=0.7) +
  labs(title = 'Distribution of AQI Values Across Cities', x = 'AQI Value', y = 'Frequency') +
  theme_minimal()

print(histogram)
```

3. Create a Density Plot of the Lung Cancer Mortality Rate. 
```{r}
mortality_rate <- ggplot(lung_cancer_data, aes(x = Mortality_Rate)) +
  geom_density(fill="blue", alpha=0.5) +
  labs(title = 'Density Plot of Lung Cancer Mortality Rate', x = 'Mortality Rate')

print(mortality_rate)
```

4. Create a Scatter Plot by generating 100 random values from both the normal and logistic
distributions. The points should be brown and use theme_solarized with argument light
set to false.
```{r}
normal_values <- rnorm(100, mean = 0, sd = 1)

logistic_values <- rlogis(100, location = 0, scale = 1)

data <- data.frame(Normal = normal_values, Logistic = logistic_values)

scatter_plot <- ggplot(data, aes(x = Normal, y = Logistic)) +
  geom_point(color = "brown", size = 2.5) +
  labs(title = "Scatter Plot of Normal vs. Logistic Random Values",
       x = "Values from Normal Distribution",
       y = "Values from Logistic Distribution") +
  theme_minimal()

print(scatter_plot)
```

### Part 4

2. Use the gpplot2 package for this graph. (Hint: Aggregate the data then merge the two
datasets. Use only the necessary columns.)

```{r}
pm25_by_country <- air_pollution_data %>%
  group_by(Country) %>%
  summarize(PM25_AQI_Value = mean(PM2.5_AQI_Value, na.rm = TRUE))

lung_cancer_deaths <- lung_cancer_data %>%
  group_by(Country) %>%
  summarize(Annual_Lung_Cancer_Deaths = sum(Annual_Lung_Cancer_Deaths, na.rm = TRUE))

merged_data <- inner_join(pm25_by_country, lung_cancer_deaths, by = "Country")

top_countries <- merged_data %>%
  arrange(desc(Annual_Lung_Cancer_Deaths)) %>%
  slice(1:3)

ggplot(merged_data, aes(x = PM25_AQI_Value, y = Annual_Lung_Cancer_Deaths, color = Country)) +
  geom_point(size = 3, alpha = 0.7) +
  geom_text(data = top_countries, aes(label = Country), color = "black", vjust = 0.5, hjust = 0.5, size = 5, fontface = "bold") + 
  labs(title = "PM2.5 AQI vs. Annual Lung Cancer Deaths",
       x = "PM2.5 AQI Value",
       y = "Annual Lung Cancer Deaths") +
  theme_minimal() +
  theme(plot.title = element_text(color = "darkred", size = 14, face = "bold"),
        legend.position = "right")
```

3. Use the ggplot2 package for this graph. (Hint: use geom_jitter since y axis contains
categorical data, also use the following colors: #5469f1 , #d554f1)
```{r}
# Remove the Stage with NA values and the values where Years of Smoking is 0
lung_cancer_data <- lung_cancer_data %>%
  select(Years_of_Smoking, Cancer_Stage, Gender) %>%
  filter(
    !is.na(Cancer_Stage) & Cancer_Stage %in% c("Stage 1", "Stage 2", "Stage 3", "Stage 4"),
    Years_of_Smoking > 0
  ) %>%
  mutate(
    Cancer_Stage = factor(Cancer_Stage, levels = c("Stage 1", "Stage 2", "Stage 3", "Stage 4"))
  )

ggplot(lung_cancer_data, aes(x = Years_of_Smoking, y = Cancer_Stage, color = Gender, shape = Gender)) +
  geom_jitter(width = 0.3, alpha = 0.6) +
  scale_color_manual(values = c("Female" = "#d554f1", "Male" = "#5469f1")) + 
  scale_shape_manual(values = c("Female" = 17, "Male" = 16)) +
  facet_wrap(~ Gender) +
  labs(
    title = "Lung Cancer Stage vs. Smoking Years",
    subtitle = "Comparison by Gender",
    x = "Years of Smoking",
    y = "Cancer Stage",
    color = "Gender",
    shape = "Gender"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12, face = "italic"),
    legend.position = "bottom"
  )
```

4. Use the ggplot2 package for this graph. (Hint: use scale_fill_viridis_d(option = "plasma"
to get the same colors)
```{r}
selected_countries <- c("Brazil", "Germany", "India", "Italy", "Russian Federation", "United States of America")

filtered_data <- air_pollution_data %>%
  filter(Country %in% selected_countries)

ggplot(filtered_data, aes(x = PM2.5_AQI_Value, fill = Country)) +
  geom_histogram(binwidth = 10, color = "black") +
  scale_fill_viridis_d(option = "plasma") +
  facet_wrap(~ Country, scales = "free_y") +
  labs(
    title = "PM2.5 AQI Distribution Across Countries",
    subtitle = "Comparison of Air Pollution Levels",
    x = "PM2.5 AQI Value",
    y = "Frequency",
    fill = "Country"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12, face = "italic"),
    legend.position = "bottom"
  )
```